---
title: "Statistical Inference on Tooth Growth Dataset"
author: "Trent Parkinson"
date: "December 26, 2017"
output:
  pdf_document
header-includes:
    - \usepackage{caption}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\captionsetup[table]{labelformat=empty}

## Overview

The project provides a basic analysis of the `ToothGrowth` data in the R datasets package. The data gives the length of the odontoblast (teeth) in each of 10 guinea pigs at three different dosage levels (0.5, 1, and 2 mg) with two supplements (Vitamin C and Orange Juice). The following will occur;

- Loads the ToothGrowth data and perform some basic exploratory data analyses.
- Provide a basic summary of the data.
- Use confidence intervals and t-tests to compare tooth growth by `supp` and `dose`.
- States conclusions and any assumptions made.

## Setting up environment

Necessary libraries for loading, manipulating, and plotting. Reading the `ToothGrowth` dataset into a `data.table`.
\vspace{4pt}
```{r, message=FALSE}

library(data.table)
library(ggplot2)
library(gridExtra)
library(dplyr)
library(printr)

data("ToothGrowth")
tooth_growth <- data.table(ToothGrowth)
```

## Data Structure

```{r}

str(tooth_growth)
head(tooth_growth)
```

## Brief Summary

\vspace{4pt}
```{r}

summary(tooth_growth)
```

## Visualizations

Setting the `dose` variable as factors to make the plotting and t-tests easier. Plotting a grid of box plots for all the different combinations of `len` as it depends on `supp` and `dose`.
\vspace{4pt}
```{r}

tooth_growth$dose <- as.factor(tooth_growth$dose)

plot1 <- ggplot(data = tooth_growth, aes(x = dose, y = len)) + 
    geom_boxplot(aes(fill = dose)) + 
    facet_grid(.~supp) +
    theme(legend.position = "None") +
    xlab(" ") + ylab("Length")

plot2 <- ggplot(data = tooth_growth, aes(x = supp, y = len)) +
    geom_boxplot(aes(fill = supp)) +
    facet_grid(.~dose) +
    theme(legend.position = "None") +
    xlab(" ") + ylab(" ")

plot3 <- ggplot(data = tooth_growth, aes(x = supp, y = len)) +
    geom_boxplot(aes(fill = supp)) +
    theme(legend.position = "None") +
    xlab("Supplement") + ylab(" ")

plot4 <- ggplot(data = tooth_growth, aes(x = dose, y = len)) +
    geom_boxplot(aes(fill = dose)) +
    theme(legend.position = "None") +
    xlab("Dosage (mg)") + ylab("Length")

grid.arrange(plot1, plot2, plot4, plot3, nrow = 2, ncol = 2)
```

## Analysis

The goal of the project is to analyze the effect of the different supplements (Vitamin C and Orange Juice) at all of the dose levels (0.5, 1, 2 mg). The following is the code used to perform all the different combinations of t-tests. All of the test were performed using a $95\%$ confidence level, with unequal variances assumed.
\vspace{4pt}
```{r}

OJvsVC <- t.test(len ~ supp, data = tooth_growth)

my.t.test <- function(fac1,fac2) {
    t.test(tooth_growth$len[tooth_growth$dose == fac1],
           tooth_growth$len[tooth_growth$dose == fac2])
}

supp.t.test <- function(supp1,dose1,supp2,dose2) {
    sub_tooth <- filter(tooth_growth, (dose == as.numeric(dose1) & supp == supp1 ) |
                            dose == as.numeric(dose2) & supp == supp2)
    t.test(len ~ supp, data=sub_tooth)
}

same_supp.t.test <- function(supp1,dose1,dose2) {
    sub_tooth <- filter(tooth_growth, (dose == as.numeric(dose1) & supp == supp1 ) |
                            dose == as.numeric(dose2) & supp == supp1)
    t.test(len ~ dose, data=sub_tooth)
}

my_tests <- mapply(FUN = my.t.test,c(.5,.5,1.0),c(1.0,2.0,2.0),SIMPLIFY = FALSE)

d1 <- c(0.5,1.0,2.0,0.5,0.5,1.0,2.0,1.0,2.0)
d2 <- c(0.5,0.5,0.5,1.0,2.0,1.0,1.0,2.0,2.0)
supp_tests <- mapply(FUN = supp.t.test,rep("VC",9),d1,rep("OJ",9),d2,SIMPLIFY = FALSE)

d3 <- rep(c("OJ","VC"),each = 3)
d4 <- rep(c(1.0,2.0,2.0), times = 2)
d5 <- rep(c(0.5,0.5,1.0), times = 2)
same_supp_tests <- mapply(FUN = same_supp.t.test,d3,d4,d5,SIMPLIFY = FALSE)
```

### Results for t-test

The following code takes all of the t-test results from the previous section and places them in a `data.frame` for easier readability, and quicker viewing. Note that all values have been rounded to four decimal places.
\vspace{8pt}
```{r,tab.cap = NULL}

t_table <- list(OJvsVC)
t_table <- append(t_table,my_tests)
t_table <- append(t_table,same_supp_tests)
t_table <- append(t_table,supp_tests)

t.stat = c(); df = c(); lower.CL = c(); upper.CL = c()
p.value = c(); mean.A = c(); mean.B = c()

rnames <- c("VC vs OJ",
            "1.0 vs 0.5","2.0 vs 0.5","2.0 vs 1.0",
            "OJ-1.0 vs 0.5","OJ-2.0 vs 0.5","OJ-2.0 vs 1.0",
            "VC-1.0 vs 0.5","VC-2.0 vs 0.5","VC-2.0 vs 1.0",
            "VC-0.5 vs OJ-0.5","VC-1.0 vs OJ-0.5","VC-2.0 vs OJ-0.5",
            "VC-0.5 vs OJ-1.0","VC-0.5 vs OJ-2.0","VC-1.0 vs OJ-1.0",
            "VC-2.0 vs OJ-1.0","VC-1.0 vs OJ-2.0","VC-2.0 vs OJ-2.0")

for (i in 1:19) {
    t.stat <- append(t.stat,c(t_table[[i]]$statistic))
    df <- append(df,c(t_table[[i]]$parameter))
    lower.CL <- append(lower.CL,c(t_table[[i]]$conf.int[1]))
    upper.CL <- append(upper.CL,c(t_table[[i]]$conf.int[2]))
    p.value <- append(p.value,c(t_table[[i]]$p.value))
    mean.B <- append(mean.B,c(t_table[[i]]$estimate[1]))
    mean.A <- append(mean.A,c(t_table[[i]]$estimate[2]))
}

mean.diff <- mean.A - mean.B

t_results <- data.frame("t.stat" = t.stat, "df" = df,
                      "lower.CL" = lower.CL, "upper.CL" = upper.CL,
                      "p.value" = p.value, "mean.A" = mean.A,
                      "mean.B" = mean.B, "mean.diff" = mean.diff,
                      row.names = rnames)

knitr::kable(round(t_results,4), caption = "Welch Two Sample t-test Results")
```

### Conclusions for t-test

Therefore the tests failed to reject the null-hypothesis (difference in means is equal to zero) in three of the cases. Since all test were held on a $95\%$ confidence level, any values with a p-value greater than $0.05$ or the confidence intervals contains the value zero will fail to reject. The results are,

- `VC vs OJ`, p-value of $0.0606$
- `VC-2.0 vs OJ-1.0`, p-value of $0.0965$
- `VC-2.0 vs OJ-2.0`, p-value of $0.9639$

## Final Conclusions

Based on the analysis performed we can conclude that as `dose` levels increase the tooth growth also increases. We can also conclude at the lower dose levels there is a statistical difference in the means of the 0.5 and 1.0 mg dose levels for `OJ` and `VC` with `OJ` being more effective on tooth growth, but at 2.0 mg `OJ` and `VC` has no significant difference. 

Further study could be conducted to see if higher dosages (above 2.0 mg) would be more effective for `VC` or `OJ`. The mean of `OJ` seems to be decreasing at a higher rate than `VC` and can be seen in the difference of means for the dose levels. `VC` may be more effective than `OJ` at higher levels.

These assumptions are based on the following:

- The sample is representative of the population
- Independent variables were randomly assigned
- The distribution of the means is normal

